---
title: "Phyloseq"
author: "Alejandra"
date: "2024-04-18"
output: pdf_document
---

#load required packages

```{r}
library(phyloseq)
library(dplyr)
library(BiMiCo)
library(ggplot2)
library(devtools)
library(MicEco)
library(vegan)
```

#load taxa and seqtab.nochim

```{r}
load("RData/taxa.RData")
load("RData/seqtab.nochim.RData")
```

#import metadata

```{r}
metadata <- read.csv("sample-metadata.csv", header=TRUE, row.names = 1)
```

#create phylseq object

```{r}
#make sure the seqtab.nochim and taxa objects are loaded
physeq <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), sample_data(metadata),tax_table(taxa))
physeq
```

#transform sample counts

```{r}
#convert from raw to abundance so its easier to comapre
physeq <- transform_sample_counts(physeq, function(abund) 1*(abund>0))

#visualize to data
physeq
```

#remove the sequence itselt and replace with ASV

```{r}
##this allows it to be easier to read, replaces the raw data
dna <- Biostrings::DNAStringSet(taxa_names(physeq))
names(dna) <- taxa_names(physeq)
physeq <- merge_phyloseq(physeq, dna)
taxa_names(physeq) <- paste0("ASV", seq(ntaxa(physeq)))
physeq
```

#remove mitochondria and phloroplast mathces, remove all non bacterial sequences

```{r}
#stictly use bacteria 16S rRNA, 
physeq <- physeq %>% subset_taxa( Family!= "Mitochondria" | is.na(Family) & Order!="Chloroplast" | is.na(Order))
physeq
```

#remove all non bacterial sequences

```{r}
physeq<-rm_nonbac(physeq)
physeq
```

#save physeq objects to load later

```{r}
save(physeq, file= "RData/physeq.RData")
```

#load physeq objects to start here

```{r}
load("RData/physeq.RData")
```

#plot bar grpah based on phylum

```{r}
plot_bar(physeq, fill = "Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="fill")
```

#create a barplot of relative abundance

```{r}
#convert to relative abundance
physeq_relabund <- transform_sample_counts(physeq, function(x) x / sum(x))

#barplot
plot_bar(physeq_relabund, fill = "Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="fill") + facet_wrap(~body.site, scales= "free")
##can change based on the column name in metadata in facet_wrap(~columnName)
```

#plot alpha diversity based on body site

```{r}
plot_richness(physeq, x="body.site", color= "subject", measures=c("Observed", "Simpson", "Shannon"))
##Simpson(less sensitive, will be more custered together) and Shannon(more sensitive to rare taxa) take into consideration relative abundance each species.
```

#pplot alpha diversity based on reported.antibiotic.usage

```{r}
plot_richness(physeq, x="reported.antibiotic.usage", color= "body.site", measures=c("Observed", "Simpson", "Shannon"))
```

#plot alpha diversity based on subject

```{r}
plot_richness(physeq, x="subject", color= "body.site", measures=c("Observed", "Simpson", "Shannon"))
```

#test for normailoty

```{r}
alpha <- estimate_richness(physeq, measures=c("Observed", "Simpson", "Shannon"))

#alternative = the data is normal distributed
#null = the data is not normal distributed

#Shapiro-wilk - used when it has fewer than 50 samples 
observed <- shapiro.test(alpha$Observed)
shannon <- shapiro.test(alpha$Shannon)
simpson <- shapiro.test(alpha$Simpson)

#print
print(observed)
print(shannon)
print(simpson)
```

#create data frames for statistical analyses

```{r}
#extract sample information from the physeq object
samples <- sample_data(physeq)

#if sample is a phyloseq sample_Data, convert it to a data frame
if (class(samples) == "sample_data") {
  samples <- data.frame(sample_data(samples))
  }

#add a column to alpha with sample names
alpha$sample <- rownames(alpha)

#merge alpha diversity data and sample(meta) data 
alpha <- merge(alpha, samples, by = "sample")
```

#perform statistics based on subject

```{r}
#perform t/wilcox tests for each biodiversity index
test_observed <- wilcox.test(Observed ~ subject, data = alpha)
test_simpson <- wilcox.test(Simpson ~ subject, data = alpha)
test_shannon <- t.test(Shannon ~ subject, data = alpha)

#print results
print(test_observed)
print(test_simpson)
print(test_shannon)
```
###to change text, select chunk > edit > find >click in selection > replace these tabs with what you ant to change; find - replace > all

#perform statistics based on reported.anitbiotic.usage
```{r}
#perform t/wilcox tests for each biodiversity index
test_observed <- wilcox.test(Observed ~ reported.antibiotic.usage, data = alpha)
test_simpson <- wilcox.test(Simpson ~ reported.antibiotic.usage, data = alpha)
test_shannon <- t.test(Shannon ~ reported.antibiotic.usage, data = alpha)

#print results
print(test_observed)
print(test_simpson)
print(test_shannon)
```
#test for body site
```{r}
kruskal.test(Simpson ~ body.site, data=alpha)
pairwise.wilcox.test(alpha$Simpson, alpha$body.site, p.adjust.method = "holm")
kruskal.test(Observed ~ body.site, data=alpha)
pairwise.wilcox.test(alpha$Observed, alpha$body.site, p.adjust.method = "holm")
shannonanova <- aov(Shannon ~ body.site, data=alpha)
summary(shannonanova)
TukeyHSD(shannonanova)
```

#load physeq objects to start here

```{r}
load("RData/physeq.RData")
```

#remove taxa with relative abundance <0.05%
```{r}
minTotRelAbun = .00005
x = taxa_sums(physeq)
keepTaxa = (x / sum(x)) > minTotRelAbun
physeqprune = prune_taxa(keepTaxa, physeq)
physeqprune
```
#nuber of shared ASVs body.site (found in 25% or more)
```{r}
#create a ven diagram showing the different categories and what they share
bodysite=ps_venn(
physeqprune,
"body.site", 
fraction = .25, 
weight = FALSE, 
relative = TRUE, 
plot = TRUE) 
bodysite
```
#bray curtis caculation, 0; exactly the same, 1; very diverse 
```{r message=FALSE}
set.seed(666)
dist = phyloseq::distance(physeqprune, method="bray", weighted=TRUE)
ordination = ordinate(physeqprune, method="NMDS", distance=dist)
```

#bray curtis subject plot
```{r}
braysubject=plot_ordination(physeqprune, ordination, color="subject") + 
  theme_classic() +
  theme(strip.background = element_blank()) + stat_ellipse(aes(group=subject))
braysubject
```

bray curtis body site plot
```{r}
braybodysite=plot_ordination(physeqprune, ordination, color="body.site") + 
  theme_classic() +
  theme(strip.background = element_blank()) + stat_ellipse(aes(group=body.site))
braybodysite
```

bray curtis antibiotic usage plot
bray curtis body site plot
```{r}
brayabusage=plot_ordination(physeqprune, ordination, color="reported.antibiotic.usage") + 
  theme_classic() +
  theme(strip.background = element_blank()) + stat_ellipse(aes(group=reported.antibiotic.usage))
brayabusage
```

#bray curits subject statistics
```{r}
adonis2(dist ~ sample_data(physeqprune)$subject)
```

#bray curtis body site statistics
```{r}
adonis2(dist ~ sample_data(physeqprune)$body.site)
ps.disper<-betadisper(dist, sample_data(physeqprune)$body.site)
permutest(ps.disper, pair=TRUE)
```

#bray curtis antibiotic usage statistics
```{r}
adonis2(dist ~ sample_data(physeqprune)$reported.antibiotic.usage)
```

#export tiff with 300dpi
```{r}
ggsave(
  filename= "Fig1.tiff",
  plot = braybodysite,
  width = 200,
  height = 200,
  units = c("mm"),
  dpi = 300
```
  
)